Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-03-12 young_pp_discs.tex
all minor changes done and stared on apendix reworking

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. \citealt{Krasnopolsky2011,Machida2011,Braiding2012,Joos2013}). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,au and masses as large as 10\% the original core mass (e.g. \citealt{Joos2012}), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well to filter out the extended emission tracing envelope material. \smallskip

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/rotational_velocities.eps}
 \caption{Azimuthaly averaged rotational velocity in the disc mid-plane with best fit curve.}
 \label{velocity}
\end{figure}

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure is derived by the hydrodynamic model of \citet{Boley2007} and  \citet{Boley2009}. The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating Keplerian disc (figure \ref{velocity}) featuring prominent spiral arms, rotating around a 1$\,$M$_\odot$ mass protostar (which is not included in the model). H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$au in $x$, $y$ and $z$. The gas/dust mass ratio was assumed to be 1/100 throughout the disc and the envelope. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history. In the dense conditions of the protoplanetary disc could be encouraging the dust grains to agglomerate and increase the size of the ice mantle through freeze out of volitiles. This could lead to changes in the dust opacity, however we have not attempted to include these effects in our model.\smallskip

@@

Figure \ref{spectra} shows spectra from the lines simulated, one from the centre of the image and one on each side of the disc along the Y=0 axis. As with all the CASA simulated images, these spectra have a total bandwidth of 938$\,$MHz, giving velocity resolutions of between 418 and 527 ms$^{-1}$. The most notable thing about these spectral features is that they are not symmetric about the disk centre. For example, C$^{17}$O(3$\rightarrow$2) has a significantly stronger blueshifted absorption, due to the presence of a spiral arm along the same line of sight. In the OCS lines, the two sides are not moving with equal velocities and the line widths and shapes are markedly different. These can be attributed to the non-axisymmetric nature of the model.

>>>>

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. \citealt{Krasnopolsky2011,Machida2011,Braiding2012,Joos2013}). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,au and masses as large as 10\% the original core mass (e.g. \citealt{Joos2012,Hayfield2011,Machida2011}), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well to filter out the extended emission tracing envelope material. \smallskip

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/rotational_velocities.eps}
 \caption{Azimuthaly averaged rotational velocity in the disc mid-plane. The best fit curve follows the equation $\frac{31.5}{\sqrt{x+1.68}}+0.518$ .}
 \label{velocity}
\end{figure}

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure used is from the same hydrodynamics simulation described in I2011, which is a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citet{Boley2007}.  The star is allowed to move freely in response to disc torques, and radiative cooling is included \citet{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H2 number densities in the disc ranging from 10$^4$-10$^13$ cm$^{-3}$ and temperatures from 30-400 K (figure \ref{rhoT}). For the non-LTE 3D radiative transfer modeling presented here, we interpolate the I2011 simulation output on to a 256$\times$256$\times$64 Cartesian grid with spatial resolution of 0.5$\,$au in x, y, and z and use this to sample the unstructured radiative transfer grid (see section \ref{subsec:gridding}).  As in the hydrodynamics simulation, the dust and gas temperatures are assumed to be in equilibrium within the disc and a gas to dust mass ratio of 1/100 is used throught the entirity of the model. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history. In the dense conditions of the protoplanetary disc could be encouraging the dust grains to agglomerate and increase the size of the ice mantle through freeze out of volitiles. This could lead to changes in the dust opacity, however we have not attempted to include these effects in our model.\smallskip

@@

Figure \ref{spectra} shows spectra from the lines simulated, one from the centre of the image and one on each side of the disc along the Y=0 axis. As with all the CASA simulated images, these spectra have a total bandwidth of 938$\,$MHz, giving velocity resolutions of between 418 and 527 ms$^{-1}$. The most notable thing about these spectral features is that they are not symmetric about the disc centre. For example, C$^{17}$O(3$\rightarrow$2) has a significantly stronger blueshifted absorption, due to the presence of a spiral arm along the same line of sight. In the OCS lines, the two sides are not moving with equal velocities and the line widths and shapes are markedly different. These can be attributed to the non-axisymmetric nature of the model.
